An investigation of factors affecting inferences from and reliability of state-space age-structured assessment models

### 2025-12-04 Project_0/manuscript/manuscript_R1.tex
saving revisions for project_0 manuscript

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{30 September, 2025}

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within International Council for
the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
2021). State-space models latent population characteristics as
statistical time series with periodic observations that also may have
error due to sampling or other sources of measurement error. Traditional
assessment models may use state-space approaches to account for temporal
variability in population characteristics (Legault and Restrepo 1999;
Methot and Wetzel 2013), but these models treat the annual parameters as
penalized fixed effects parameters where the variance parameters
controlling the penalties are assumed known (Thorson and Minto 2015).
Modern state-space models can estimate the annually varying parameters
as random effects with variance parameters estimated using maximum
marginal likelihood or corresponding Bayesian approaches. These latter
approaches are considered best practice and a recommended for the next
generation of stock assessment models (Hoyle et al. 2022; Punt 2023).

@@

As state-space models provide greater flexibility by allowing multiple
processes to vary as random effects (Nielsen and Berg 2014; Aeberhard et
al. 2018; Stock et al. 2021), one of the most immediate questions
regards the implications of mis-specification among alternative sources
of process error. Incorrect treatment of population attributes as
temporally varying (Trijoulet et al. 2020; Liljestrand et al. 2024)
could lead to misidentification of stock status and biased population
estimates, ultimately impacting fisheries management decisions (Legault
and Palmer 2016; Szuwalski et al. 2018; Cronin-Fine and Punt 2021).
Furthermore, biological, fishery, and observational processes are often
confounded in catch-at-age data, which may adversely affect ability to
distinguish between true process variability and observational error (Li
et al. In review; Punt et al. 2014; Stewart and Monnahan 2017;
Cronin-Fine and Punt 2021; Fisch et al. 2023).

@@

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model were not
thoroughly examined or tested within a simulation-estimation framework.
Second, previous studies relied on operating models conditioned on
specific fisheries, limiting their generalizability (Li et al. In
review; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history have not been fully
isolated in simulation study designs, making it challenging to
disentangle the interplay between process and observation error
magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
modeling stock-recruit relationships (SRRs) as mechanistic drivers of
population dynamics is promising (Fleischman et al. 2013; Pontavice et
al. 2022), but reliability of inferences within integrated state-space
age-structured models has not been evaluated. Evidence from other
studies suggests that when both process and observation errors are
unknown, estimating density dependence parameters becomes highly
uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
(2008) demonstrated that stronger density dependence becomes
increasingly difficult to estimate in the presence of observation error.
Therefore, it is crucial to assess whether density dependence mechanisms
can be estimated with sufficient precision for use in fisheries
management (Auger-Méthé et al. 2016). Finally, although the importance
of autocorrelation in process errors is recognized, investigations of
the ability to distinguish state-space assessment models with and
without autocorrelation and whether such misspecification is detrimental
to estimation of important population metrics are lacking (Johnson et
al. 2016; Xu et al. 2019).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments ({Auger-Méthé et al.} 2021). We evaluate
whether convergence and Akaike Information Criterion (AIC) can correctly
determine the source of process error and the existence of a SRR. We
also evaluate when retrospective patterns occur and the degree of bias
in outputs of the assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Miller and Stock 2020; Stock and Miller 2021).
WHAM is an R package freely available via a github repository and is
built on the Template Model Builder package (Kristensen et al. 2016).
For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
and EMs for closed loop simulations evaluating index-based assessment
methods (Legault et al. 2023) and is currently used or accepted for use
in management of numerous NEUS fish stocks (e.g., NEFSC 2022a, 2022b;
NEFSC 2024).

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects were assumed:
recruitment (R, assumed present in all models), apparent survival
(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
survival because on logscale the random effects (\(\epsilon_{a,y}\)) are
additive to the total mortality (F+M) between numbers at age, thus they
modify the survival term. However, as Stock and Miller (2021) note,
these random effects can be due to events other than mortality, such as
immigration, emigration, missreported catch, and other sources of
misspecification. For each OM, assumptions about the magnitude of the
variance of process errors and observations are required and the values
we used were based on a review of the range of estimates from Northeast
Unite States (NEUS) assessments using WHAM.

In total, we configured 72 OMs with alternative assumptions about the
source and magnitude of process errors, magnitude of observation error
in indices and age composition data, and contrast in fishing pressure
over time. We fitted 20 EMs to observations generated from each of 100
simulations where process errors were also simulated. Each EM differed
in assumptions about the source of process errors, whether natural
mortality (or the median for models with process error in natural
mortality) was estimated, and whether a Beverton-Holt SRR was estimated
within the EM. Details of each of the OMs and EMs are described below.

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for OMs and EMs to
simplify interpretation of the study results (Li et al. In review). All
code we used to perform the simulation study and summarize results can
be found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

The population consists of 10 age classes, ages 1 to 10+, with the last
being a plus group that accumulates ages 10 and older. We assume
spawning occurs annually 1/4 of the way through the year. The maturity
at age was a logistic curve with \(a_{50}\) = 2.89 and slope = 0.88
(Figure \ref{om_inputs_fig}, top left).

@@

We assumed a Beverton-Holt SRR with constant pre-recruit mortality
parameters for all OMs. All biological inputs to calculations of
spawning biomass per recruit (i.e., weight, maturity, and natural
mortality at age) are constant in the apparent survival (R+S)
selectivity (R+Sel), and survey catchability (R+q) process error OMs.
Therefore, steepness and unfished recruitment are also constant over the
time period for those OMs (Miller and Brooks 2021). We specified
unfished recruitment equal to \(e^{10}\) and
\(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69 and
\(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
parameterization \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] (Figure \ref{om_inputs_fig}, bottom right). We assumed a value of 0.2
for the natural mortality rate in OMs without process errors on natural
mortaliyt and for the median rate for OMs with process errors on natural
mortality.

We used two fishing scenarios for OMs. In the first scenario, the stock
experiences overfishing at 2.5\Fmsy for the first 20 years followed by
fishing at \Fmsy for the last 20 years (denoted
\(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is
fished at \Fmsy for the entire time period (40 years). The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
the approach in Legault et al. (2023).

We specified initial population abundance at age at the equilibrium
distribution that corresponds to fishing at either
\(F = 2.5\times \Fmsy\) or \(F = \Fmsy\). This implies that, for a
deterministic model, the abundance at age would not change from the
first year to the next.

For OMs with time-varying random effects for M , steepness is not
constant. However, we used the same \(a\) and \(b\) parameters as other
OMs, which equates to a steepness and R0 at the median of the time
series process for M. For OMs with time-varying random effects for
fishery selectivity, \Fmsy is also not constant, but since we use the
same F history as other OMs, this corresponds to \Fmsy at the mean
selectivity parameters.

@@

We assumed a single fleet operating year round for catch observations
with logistic selectivity for the fleet (\(a_{50} = 5\) and slope = 1;
Figure \ref{om_inputs_fig}, bottom left). This selectivity was used to
define \Fmsy for the Beverton-Holt SRR parameters above. We assumed a
logistic-normal distribution with no correlation on the multivariate
normal scale for the age-composition observations for the fleet.

@@

Two time series of fishery-independent surveys in numbers are generated
for the entire 40 year period with one occurring in the spring (0.25 of
each year) and one in the fall (0.75 of each year). Catchability of both
surveys are assumed to be 0.1. Like the fishing fleet, we assumed
logistic selectivity for both indices (\(a_{50} = 5\) and slope = 1) and
a logistic-normal distribution with no correlation on the multivariate
normal scale for the age-composition observations.

@@

For operating models with random effects on recruitment and(or) apparent
survival (R, R+S), we assumed marginal standard deviations for
recruitment of \(\sigma_R \in \{0.5,1.5\}\) and marginal standard
deviations for older age classes of \(\sigma_{2+} \in \{0,0.25, 0.5\}\).
The full factorial combination of these process error assumptions (2x3
levels) and scenarios for fishing history (2 levels) and observation
error (2 levels) scenarios described above results in 24 different R
(\(\sigma_{2+} = 0\)) and R+S operating models (Table
\ref{naa_om_table}).

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit (e.g., Carvalho et al.
2021; Kapur et al. 2025), but given the importance of estimates of
uncertainty when using assessment models in management, we estimated
probability of convergence as measured by occurrence of a
positive-definite hessian matrix at the optimized negative
log-likelihood that could be inverted (i.e., providing hessian-based
standard error estimates). We also provide results in the Supplementary
Materials for convergence defined by the maximum absolute gradient
\textless{} \(1^{-6}\) and the maximum of the absolute gradient values
for all fits of a given EM to all simulated data sets from a given OM
that produced hessian-based standard errors for all estimated fixed
effects. This provides an indication of how poor the calculated
gradients can be, but still presumably converged adequately enough for
parameter inferences.

@@

We investigated the reliability of AIC-based model selection for two
purposes. First, we analyzed selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
simulated data set, we compared AIC for EMs with different process error
assumption conditional on whether median natural mortality rate and the
Beverton-Holt SRR were estimated. We tabulated the models providing the
lowest AIC across simulated data sets. Second, we analyzed AIC-based
selection between EMs with and without the Beverton-Holt SRR assumed.
Contrast in fishing pressure and time series with recruitment at low
stock size has been shown to improve estimation of SRR parameters
(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
inspections indicated generally poor performance of AIC in determining
the Beverton-Holt model for a given set of OM factors (including
contrast in fishing pressure). Therefore, we also considered the effect
of of the log-standard deviation of the true log(SSB) (similar to the
log of the coefficient of variation for SSB) on model selection since
simulations with realized SSB producing low and high recruitment would
have larger variation in realized SSB.

@@

We also investigated bias in estimation of various model attributes as a
measure of reliability. For a given model attribute we calculated the
relative error \begin{equation}\label{relerror}
\text{RE}\left(\theta_i\right) = \frac{\widehat \theta_i - \theta_i}{\theta_i}
\end{equation} from fitting a given estimating model to simulated data
set \(i\) configured for a given OM where \(\widehat \theta_i\) and
\(\theta_i\) are the estimated and true values for simulation \(i\). We
analyzed simulation results for estimates of terminal year estimates of
SSB and recruitment, Beverton-Holt SRR parameters (\(a\) and \(b\)), and
median natural mortality rate.

@@

Finally, we investigated presence of retrospective patterns in fitted
models as a measure of reliability. We calculated Mohn's \(\rho\) for
SSB, fishing mortality (averaged over all age classes), and recruitment
for each EM fit to each OM simulated data set (Mohn 1999). We fit
\(P = 7\) peels to each simulated data set and calculated Mohn's
\(\rho\) for a given attribute \(\theta\) as
\begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation} where \(\widehat \theta_{i,j}\) is the estimate for
attribute \(\theta\) in year \(i\) from a model fit using data up to
year \(j\).

@@

The measures of central tendency and variability of observed values for
specific OM and EM attributes (e.g., low or high observation error) that
we described above can indicate scenarios that provide better or poorer
reliability. However, the OM and EM attributes that we investigated are
numerous, so we used two methods to summarize the most important factors
for differences in results. The first method was fitting regression
models with the response being each of the measures of reliability
described above and predictor variables were defined based on OM and EM
characteristics (e.g., MacKinnon et al. 1995; Wang et al. 2017; Harwell
et al. 2018). For the binary indicators of convergence and AIC-based
selection of a SRR, we performed logistic regressions. For indicators of
AIC-based selection of EM process error type (multiple categories) we
performed multinomial regressions. For other measures of reliability we
fit linear regression models to transformed responses. Because relative
errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the various
parameters are bounded below at -1, we used a transformation of these
values \begin{equation}\label{bias_regression_response}
y_i = \log\left[f\left(\widehat \theta_i,\theta_i\right)+1\right]
\end{equation} where \(f\) is either the relative error (Eq.
\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
\(i\), so that values are unbounded. For relative errors, \(y_i\) is the
log-scale error. We omitted simulation estimates equal to zero (RE =
-1). For all regressions we fit separate models with individual factors
included, with all factors combined, with including all second order
interactions, and including all third order interactions. For the
multinomial regression, we used the \verb|vglm| function from the VGAM
package (Yee 2008; Yee 2015). We tabulated percent reduction in residual
deviance for each of regression fits. We did not perform formal
statistical analyses of effects of OM and EM attributes on results
(e.g., ANOVA) because of the lack of independence of the
``observations'' that results from fitting multiple EMs to each
simulated data set.

@@

We also provide detailed results for all measures of reliability at each
combination of OM and EM attributes in the Supplementary Materials. For
confidence intervals of probability of convergence, we used the
Clopper-Pearson exact method (Clopper and Pearson 1934; Thulin 2014).
For AIC selection of process error configuration we provide estimates of
the proportions of simulations where each EM type was selected. For AIC
selection of the SRR, we provide predicted probabilities from logistic
regressions as a function of SSB variability for each OM and EM type. We
estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. We constructed 95\%
confidence intervals for the median relative bias, and Mohn's \(\rho\)
using the binomial distribution approach as in Miller and Hyun (2018)
and Stock and Miller (2021). For each EM we calculated median and 95\%
confidence intervals using the same methods as that for relative bias.

@@

For probability of convergence, EM process error assumption was the
single attribute that resulted in the largest percent reduction in
deviance (14-28\%) for all OM process error types other than R+S OMs
where the EM M assumption explained the most residual deviance
(\textgreater11\%; Table \ref{convergence_PRD_table}). However,
including interactions of OM and EM factors also provided large
reductions in residual deviance (35-47\%), suggesting successful
convergence depended on a combination OM and EM attributes.

Classification trees for each OM process error type, all had the primary
branch defined using the same attribute that provided the largest
reduction in deviance (Figure \ref{conv_class}). EMs that assumed R+S
process errors converged poorly for all OMs that were simulated with the
alternative process error assumptions (R, R+M, R+Sel, an R+q OMs). For
all trees, branches based on the OM fishing mortality history showed
better convergence when the OM included a change in fishing pressure.
Branches based on whether the Beverton-Holt SRR was assumed or not,
showed better convergence when it was not estimated and branches based
on whether the EM estimated the median natural mortality rate showed
better convergence when it was treated as known. For certain R+M and
R+Sel OMs, better convergence was also observed when there was lower
observation uncertainty.

>>>>

\title{An investigation of factors affecting inferences from and
reliability of state-space age-structured assessment models}
\author{Timothy J. Miller\textsuperscript{1,2} \and Greg
Britten\textsuperscript{3} \and Elizabeth N.
Brooks\textsuperscript{2} \and Gavin Fay\textsuperscript{4} \and Alex
Hansell\textsuperscript{2} \and Christopher M.
Legault\textsuperscript{2} \and Chengxue
Li\textsuperscript{2} \and Brandon Muffley\textsuperscript{5} \and Brian
C. Stock\textsuperscript{6} \and John Wiedenmann\textsuperscript{7}}
\date{04 December, 2025}

@@

Application of state-space models in fisheries stock assessment and
management has expanded dramatically within International Council for
the Exploration of the Sea (ICES), Canada, and the Northeast US (Nielsen
and Berg 2014; Cadigan 2016; Pedersen and Berg 2017; Stock and Miller
2021). State-space models treat latent population characteristics as
statistical time series with periodic observations that also may have
error due to sampling or other other measurement properties. Traditional
assessment models may use state-space approaches to account for temporal
variability in population characteristics (Legault and Restrepo 1999;
Methot and Wetzel 2013), but these models treat the annual parameters as
penalized fixed effect parameters where the variance parameters
controlling the penalties are assumed known (Thorson and Minto 2015).
Modern state-space models can estimate the annually varying parameters
as random effects with variance parameters estimated using maximum
marginal likelihood or corresponding Bayesian approaches. These latter
approaches are considered best practice and are recommended for the next
generation of stock assessment models (Hoyle et al. 2022; Punt 2023).

@@

As state-space models provide greater flexibility by allowing multiple
processes to vary as random effects (Nielsen and Berg 2014; Aeberhard et
al. 2018; Stock et al. 2021), one of the most immediate questions
regards the implications of mis-specification among alternative sources
of process error. Incorrect treatment of population attributes as
temporally varying (Trijoulet et al. 2020; Liljestrand et al. 2024)
could lead to misidentification of stock status and biased population
estimates, ultimately impacting fisheries management decisions (Legault
and Palmer 2016; Szuwalski et al. 2018; Cronin-Fine and Punt 2021).
Furthermore, biological, fishery, and observational processes are often
confounded in catch-at-age data, which may adversely affect the ability
to distinguish between true process variability and observational error
(Li et al. In press; Punt et al. 2014; Stewart and Monnahan 2017;
Cronin-Fine and Punt 2021; Fisch et al. 2023).

@@

Despite mounting efforts, several limitations remain. First, confounding
processes that can be treated as random effects in the model have not
been thoroughly examined or tested within a simulation-estimation
framework. Second, previous studies relied on operating models
conditioned on specific fisheries, limiting their generalizability (Li
et al. In press; Liljestrand et al. 2024). In particular, the effects of
observation error and underlying fishing history have not been fully
isolated in simulation study designs, making it challenging to
disentangle the interplay between process and observation error
magnitudes, as demonstrated in Fisch et al. (2023). Third, explicitly
modeling stock-recruit relationships (SRRs) as mechanistic drivers of
population dynamics is promising (Fleischman et al. 2013; Pontavice et
al. 2022), but reliability of inferences within integrated state-space
age-structured models has not been evaluated. Evidence from other
studies suggests that when both process and observation errors are
unknown, estimating density dependence parameters becomes highly
uncertain (Knape 2008; Polansky et al. 2009). In particular, Knape
(2008) demonstrated that stronger density dependence becomes
increasingly difficult to estimate in the presence of observation error.
Therefore, it is crucial to assess whether density dependence mechanisms
can be estimated with sufficient precision for use in fisheries
management (Auger-Méthé et al. 2016). Finally, although the importance
of autocorrelation in process errors is recognized, investigations of
the ability to distinguish state-space assessment models with and
without autocorrelation and whether such misspecification is detrimental
to estimation of important population metrics are lacking (Johnson et
al. 2016; Xu et al. 2019).

In the present study, we conduct a simulation study with operating
models (OMs) varying by degree of observation error, source and
variability of process error, and fishing history. The simulations from
these OMs are fitted with estimation models (EMs) that make alternative
assumptions for sources of process error, whether a SRR was estimated,
and whether natural mortality is estimated. Given the confounding nature
of process errors, developing diagnostic tools to detect model
misspecification is of great scientific interest and could aid the next
generation of stock assessments ({Auger-Méthé et al.} 2021). We evaluate
whether OM and EM attributes affect rates of convergence and the ability
of Akaike Information Criterion (AIC) to correctly determine the source
of process error or the existence of a SRR. We also evaluate when
retrospective patterns occur and the degree of bias in outputs of the
assessment model that are important for management.

@@

We used the Woods Hole Assessment Model (WHAM) to configure OMs and EMs
in our simulation study (Stock and Miller 2021; Miller et al. 2025).
WHAM is an R package freely available via a Github repository and is
built on the Template Model Builder package (Kristensen et al. 2016).
For this study we used
\href{https://github.com/timjmiller/wham/tree/77bbd946e4881216a439933473d1c58b21c270c3}{version
1.0.6.9000, commit 77bbd94}. WHAM has also been used to configure OMs
and EMs for closed loop simulations evaluating index-based assessment
methods (Legault et al. 2023) and is currently used or accepted for use
in management of numerous NEUS fish stocks (e.g., NEFSC 2022a, 2022b;
NEFSC 2024).

We completed a simulation study with a number of OMs that can be
categorized based on where process error random effects were assumed:
recruitment (R, assumed present in all models), apparent survival
(denoted R+S), natural mortality (R+M), fleet selectivity (R+Sel), or
index catchability (R+q). We refer to the (R+S) OMs as modeling apparent
survival because on log-scale the random effects are additive to the
total mortality (F+M) between numbers at age, thus they modify the
survival term. However, as Stock and Miller (2021) note, these random
effects can be due to events other than mortality, such as immigration,
emigration, missreported catch, and other sources of misspecification.
For each OM, assumptions about the magnitude of the variance of process
errors and observations are required and the values we used were based
on a review of the range of estimates from Northeast Unite States (NEUS)
assessments using WHAM.

In total, we configured 72 OMs with alternative assumptions about the
source and magnitude of process errors, magnitude of observation error
in indices and age composition data, and contrast in fishing pressure
over time. For each OM, we simulated 100 population time series with
process errors and, for each time series, simulated observation data
sets. For each data set, we fitted a number of EMs that differed in
assumptions about the source of process errors, whether natural
mortality (or the median for models with process error in natural
mortality) was estimated, and whether a Beverton-Holt SRR was estimated
within the EM. Details of each of the OMs and EMs are described below.

We did not use the log-normal bias-correction feature for process errors
or observations described by (Stock and Miller 2021) for OMs and EMs to
simplify interpretation of the study results (Li et al. 2025). All code
we used to perform the simulation study and summarize results can be
found at
\url{https://github.com/timjmiller/SSRTWG/tree/main/Project_0/code}.

@@

We intended the population demographics and observation types to
represent a general NEUS groundfish stock. The population consists of 10
age classes, ages 1 to 10+, with the last being a plus group that
accumulates ages 10 and older. We assume spawning occurs annually 1/4 of
the way through the year. The maturity at age was a logistic curve with
\(a_{50}\) = 2.89 and slope = 0.88 (Figure \ref{om_inputs_fig}, top
left).

@@

We assumed a Beverton-Holt SRR with constant pre-recruit mortality
parameters for all OMs. All biological inputs to calculations of
spawning biomass per recruit (i.e., weight, maturity, and natural
mortality at age) are constant in the apparent survival (R+S)
selectivity (R+Sel), and survey catchability (R+q) process error OMs.
Therefore, steepness and unfished recruitment are also constant over the
time period for those OMs (Miller and Brooks 2021). We assumed a value
of 0.2 for the natural mortality rate in OMs without process errors on
natural mortality. We specified unfished recruitment equal to \(e^{10}\)
and \(\Fmsy = F_{40\%} = 0.348\), which equates to a steepness of 0.69
and \(a=0.60\) and \(b = 2.4 \times 10^{-5}\) for the Beverton-Holt
parameterization \[
N_{1,y} = \frac{a \text{SSB}_{y-1}}{1 + b \text{SSB}_{y-1}} 
\] (Figure \ref{om_inputs_fig}, bottom right). For OMs with time-varying
random effects for natural mortality, steepness is not constant.
However, we used the same \(a\) and \(b\) parameters as other OMs, which
equates to a steepness and R0 at the median of the time series process
for natural mortality. Similarly, For OMs with time-varying random
effects for fishery selectivity, \Fmsy also varies temporally, so
equilibrium conditions for these OMs are defined for mean selectivity
parameters.

We used two fishing scenarios for OMs. In the first scenario, the stock
experiences overfishing at 2.5\Fmsy for the first 20 years followed by
fishing at \Fmsy for the last 20 years (denoted
\(2.5\Fmsy \rightarrow \Fmsy\)). In the second scenario, the stock is
fished at \Fmsy for the entire time period (40 years). The magnitude of
the overfishing assumptions is based on average estimates of overfishing
for NEUS groundfish stocks from Wiedenmann et al. (2019) and similar to
the approach in Legault et al. (2023). The second scenario represents
the ideal situation where the stock is fished at an optimal level, but
provides less contrast in stock sizes over time. We specified initial
population abundance at age at the equilibrium distribution that
corresponds to fishing at either \(F = 2.5\Fmsy\) or \(F = \Fmsy\). This
implies that, for a deterministic model, the abundance at age would not
change from the first year to the next.

@@

We assumed a single fleet operating year round for catch observations
with logistic selectivity (\(a_{50} = 5\) and slope = 1; Figure
\ref{om_inputs_fig}, bottom left). This selectivity was used to define
\Fmsy for the Beverton-Holt SRR parameters above. We assumed a
logistic-normal distribution with no correlation on the multivariate
normal scale for the corresponding annual age-composition observations.

@@

Two time series of fishery-independent surveys measured in numbers are
generated for the entire 40 year period with one occurring in the spring
(0.25 of each year) and one in the fall (0.75 of each year),
representing current bottom trawl surveys conducted in the NEUS.
Catchability of both surveys are assumed to be 0.1. Like the fishing
fleet, we assumed logistic selectivity for both indices (\(a_{50} = 5\)
and slope = 1) and a logistic-normal distribution with no correlation on
the multivariate normal scale for the annual age-composition
observations.

@@

For operating models with random effects on recruitment and(or) apparent
survival (R, R+S), we assumed marginal standard deviations for
recruitment of \(\sigma_R \in \{0.5,1.5\}\). The marginal standard
deviations for apparent survival random effects at older age classes
were \(\sigma_{2+} \in \{0,0.25, 0.5\}\). The full factorial combination
of these process error assumptions (\(2\times 3\) levels) and scenarios
for fishing history (2 levels) and observation error (2 levels)
scenarios described above results in 24 different R
(\(\sigma_{2+} = 0\)) and R+S operating models (Table
\ref{naa_om_table}).

@@

The first measure of reliability we investigated was frequency of
convergence when fitting each EM to the simulated data sets. There are
various ways to assess convergence of the fit (e.g., Carvalho et al.
2021; Kapur et al. 2025), but given the importance of estimates of
uncertainty when using assessment models in management, we estimated
probability of convergence as measured by occurrence of a
positive-definite Hessian matrix at the optimized negative
log-likelihood that could be inverted (i.e., providing Hessian-based
standard error estimates). We also provide results in the Supplementary
Materials for convergence defined by the maximum absolute gradient
\textless{} \(1^{-6}\) and the maximum of the absolute gradient values
for all fits of a given EM to all simulated data sets from a given OM
that produced Hessian-based standard errors for all estimated fixed
effects. This provides an indication of how poor the calculated
gradients can be, but still presumably converged adequately enough for
parameter inferences.

@@

We investigated the reliability of AIC-based model selection for two
purposes. First, we analyzed selection of each process error model
structure (R, R+S, R+M, R+Sel, R+q) using marginal AIC. For a given OM
simulated data set, we compared AIC for EMs with different process error
assumption conditional on whether median natural mortality rate and the
Beverton-Holt SRR were estimated. Second, we analyzed AIC-based
selection between EMs with and without the Beverton-Holt SRR assumed.
Contrast in fishing pressure and time series with recruitment at low
stock size has been shown to improve estimation of SRR parameters
(Magnusson and Hilborn 2007; Conn et al. 2010). Our preliminary
inspections indicated generally poor performance of AIC in determining
the Beverton-Holt model for a given set of OM factors (including
contrast in fishing pressure), even when the EM was configured with the
correct process error type. Therefore, we conditioned on the EMs having
the correct process error assumption and also considered the effect of
the log-standard deviation of the true log(SSB)
(\(\log \text{SD}_{\text{SSB}}\); similar to the log of the coefficient
of variation for SSB) on model selection since simulations with realized
SSB producing low and high recruitment would have larger variation in
realized SSB.

@@

We also investigated bias in estimation of various model attributes as a
measure of reliability. For a given model attribute we calculated the
relative error \begin{equation}\label{relerror}
\text{RE}\left(\theta_j\right) = \frac{\widehat \theta_j - \theta_j}{\theta_j}
\end{equation} from fitting a given estimating model to simulated data
set \(j\) configured for a given OM where \(\widehat \theta_j\) and
\(\theta_j\) are the estimated and true values for simulation \(j\). We
analyzed simulation results for estimates of terminal year estimates of
SSB and recruitment, Beverton-Holt SRR parameters (\(a\) and \(b\)), and
median natural mortality rate.

@@

Finally, we investigated presence of retrospective patterns in fitted
models as a measure of reliability. We calculated Mohn's \(\rho\) for
SSB, fishing mortality (averaged over all age classes), and recruitment
for each EM fit to each OM simulated data set (Mohn 1999). We fit
\(P = 7\) peels to each simulated data set and calculated Mohn's
\(\rho\) for a given attribute \(\theta\) as
\begin{equation}\label{mohns_rho}
\rho\left(\theta\right) = \frac{1}{P}\sum^P_{p=1} \frac{\widehat \theta_{Y-p,Y-p} - \widehat\theta_{Y-p,Y}}{\widehat\theta_{Y-p,Y}}
\end{equation} where \(Y\) is last year of the full set of observations
and \(\widehat \theta_{y,y'}\) is the estimate for attribute \(\theta\)
in year \(y\) from a model fit using data up to year \(y'\geq y\).

@@

Because the OM and EM attributes that we investigated are numerous, we
used two methods to summarize the most important factors for differences
in results within a given OM process error type. The first method was
fitting regression models with the response being each of the measures
of reliability described above and predictor variables were defined
based on OM and EM characteristics (e.g., MacKinnon et al. 1995; Wang et
al. 2017; Harwell et al. 2018). For the binary indicators of convergence
and AIC-based selection of a SRR, we performed logistic regressions. For
indicators of AIC-based selection of EM process error type (multiple
categories) we performed multinomial regressions. For other measures of
reliability we fit linear regression models to transformed responses.
Because relative errors (Eq. \ref{relerror}) and Mohn's \(\rho\) for the
various parameters are bounded below at -1, we used a transformation of
these values \begin{equation}\label{bias_regression_response}
y_j = \log\left[f\left(\widehat \theta_j,\theta_j\right)+1\right]
\end{equation} where \(f\) is either the relative error (Eq.
\ref{relerror}) or Mohn's \(\rho\) (Eq. \ref{mohns_rho}) for simulation
\(j\), so that values are unbounded. For relative errors, \(y_j\) is the
log-scale error. We omitted simulations where estimated attributes equal
to zero (RE = -1). For all regressions we fit separate models with just
individual OM and EM factors included, with all factors included, with
all second order interactions, and with all third order interactions.
For the multinomial regression, we used the \verb|vglm| function from
the VGAM package (Yee 2008; Yee 2015). We tabulated percent reduction in
residual deviance for each of regression fits. We did not perform formal
statistical analyses of effects of OM and EM attributes on results
(e.g., ANOVA) because of the lack of independence of the
``observations'' that results from fitting multiple EMs to each
simulated data set.

@@

We also provide detailed results for all measures of reliability at each
combination of OM and EM attributes in the Supplementary Materials. For
confidence intervals of probability of convergence, we used the
Clopper-Pearson exact method (Clopper and Pearson 1934; Thulin 2014).
For AIC selection of process error configuration we provide estimates of
the proportions of simulations where each EM type was selected. For AIC
selection of the SRR (a binary indicator for each simulated data set),
we fit logistic regressions and present resulting predicted
probabilities of correctly selecting the SRR as a function of SSB
variability (\(\log \text{SD}_{\text{SSB}}\) described above). We
estimated bias as the median of the relative errors across all
simulations for a given OM and EM combination. We constructed 95\%
confidence intervals for the median relative bias, and Mohn's \(\rho\)
using the binomial distribution approach (Thompson 1936) as in Miller
and Hyun (2018) and Stock and Miller (2021).

@@

For probability of convergence, EM process error assumption was the
single attribute that resulted in the largest percent reduction in
deviance (14-28\%) for all OM process error types other than R+S OMs
where the EM median natural mortality rate assumption (estimated or
known) explained the most residual deviance (\textgreater11\%; Table
\ref{convergence_PRD_table}). However, including interactions of OM and
EM factors also provided large reductions in residual deviance
(35-47\%), suggesting successful convergence depended on a combination
OM and EM attributes.

Classification trees for each OM process error type, all had the primary
branch defined using the same attribute that provided the largest
reduction in deviance (Figure \ref{conv_class}). EMs that assumed R+S
process errors converged poorly for all OMs that were simulated with the
alternative process error assumptions (R, R+M, R+Sel, an R+q OMs). For
all trees, branches based on the OM fishing mortality history showed
better convergence when the OM included a change in fishing pressure.
Branches based on whether the Beverton-Holt SRR was assumed or not,
showed better convergence when it was not estimated and branches based
on the median natural mortality rate assumption showed better
convergence when it was treated as known. For some R+M and R+Sel OMs,
better convergence was also observed when there was lower observation
uncertainty.

When convergence is defined by a gradient threshold, the primary factor
explaining deviance reduction is the same that for Hessian-based
convergence for all OM process error types, but there are some
differences in deviance reduction for secondary factors (Table
\ref{convergence_gradient_PRD_table}), and probability of convergence,
overall, was lower (Figure \ref{conv_gradient_class}). We found a wide
range of maximum absolute values of gradients for models that had
invertible Hessians (Figure \ref{hess_grad}). The largest value observed
for a given EM and OM combination was typically \(<10^{-3}\), but many
converged models had values greater than 1. For many OMs, EMs that
assumed the correct process error type and did not estimate median
natural mortality or the Beverton-Holt SRR produced the lowest gradient
values.
